---
title: "Predicting Solar Radiation (Sunlight) from Weather Variables"
subtitle: "Data from NASA via Kaggle user Andrey (dronio)"
author: "Natalie Weaver"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_notebook
---

## Document Setup

Everything we need to make this file run properly.

```{r, setup}
# load packages
library(pacman)
p_load(tidyverse, magrittr, janitor, DT, lubridate, viridis, knitr, cowplot, caret, rpart, rpart.plot)

# knitr options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  eval = TRUE,
  fig.align = "center",
  dpi = 500)

options(knitr.table.format = "html")

# text size for ggplots
font_size <- 40
```

## The Question

Can we predict the level of solar radiation (amount of sunlight) using other meteorological features?

## The Data

### Background information

This dataset originally comes from NASA for a hackathon challenge. I downloaded it as a csv from [Kaggle](https://www.kaggle.com/dronio/SolarEnergy), where it was uploaded by a user named [Andrey](https://www.kaggle.com/dronio) in 2017.

The file contains meteorological data collected by the weather station in the [HI-SEAS](https://hi-seas.org/) habitat on the Big Island of Hawaii. HI-SEAS is a Mars analog habitat operated by the University of Hawaii; it has been used by NASA for research into human behavior in conditions simulating a long-term mission to Mars. This data was collected from September through December of 2016, between Mission IV and Mission V.

The variables in this dataset are the following:

* `UNIXTime`: the date/time of the measurement in UNIX time (the number of seconds since January 1st, 1970)
* `Data`: the date of the measurement, formatted mm/dd/yyyy (should be called `date`, will change)
* `Time`: the time of the measurement, formatted hh:mm:ss
* `Radiation`: the solar radiation (amount of sunlight) in watts per square meter
* `Temperature`: the temperature in degrees Fahrenheit
* `Pressure`: the barometric pressure in inches Hg
* `Humidity`: the humidity as a percentage
* `WindDirection(Degrees)`: the direction from which the wind blows, in degrees from North going clockwise (e.g. a wind coming from due East would have a 90 degree angle)
* `Speed`: the wind speed in miles per hour
* `TimeSunRise`: the local time of the sunrise, formatted hh:mm:ss
* `TimeSunSet`: the local time of the sunset, formatted hh:mm:ss

### Exploratory Analysis

Let's read in the data and see the column specifications, head, and number of missing values.

```{r, load-data}
# read in the data using read_csv from readr in the tidyverse
# reformat the names using clean_names from janitor
data <- read_csv("SolarPrediction.csv") %>% clean_names() %>%
  rename(date = data) # rename data column to date

# get column specs (metadata created by read_csv)
data %>% spec()

# see the head of the data
data %>% head()

# are there any missing values?
data %>% is.na() %>% colSums() %>% as.data.frame()
```
We got lucky with this dataset that there are no missing values.

Let's make some visualizations to better understand the shape of the data. Here are some histograms showing the distributions of various features in our data.

```{r, visualize-data, fig.height = 7}
radiation_hist <- ggplot(data = data, aes(x = radiation)) +
  geom_histogram(fill = "#440154FF") +
  labs(
    title = "Distribution of Radiation Values",
    x = "Radiation (watts per square meter)"
  ) +
  theme(text = element_text(size = font_size))

temperature_hist <- ggplot(data = data, aes(x = temperature)) +
  geom_histogram(fill = "#404788FF") +
  scale_x_continuous(labels = function(x) paste0(x, "°")) +
  labs(
    title = "Distribution of Temperature Values",
    x = "Temperature (degrees Fahrenheit)"
  ) +
  theme(text = element_text(size = font_size))

pressure_hist <- ggplot(data = data, aes(x = pressure)) +
  geom_histogram(fill = "#287D8EFF") +
  labs(
    title = "Distribution of Pressure Values",
    x = "Pressure (inches of Hg)"
  ) +
  theme(text = element_text(size = font_size))

humidity_hist <- ggplot(data = data, aes(x = humidity)) +
  geom_histogram(fill = "#29AF7FFF") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Distribution of Humidity Values",
    x = "Humidity (percentage)"
  ) +
  theme(text = element_text(size = font_size))

direction_hist <- ggplot(data = data, aes(x = wind_direction_degrees)) +
  geom_histogram(fill = "#95D840FF") +
  scale_x_continuous(labels = function(x) paste0(x, "°")) +
  labs(
    title = "Distribution of Wind Direction Values",
    x = "Wind Direction (degrees clockwise from North)"
  ) +
  theme(text = element_text(size = font_size))

speed_hist <- ggplot(data = data, aes(x = speed)) +
  geom_histogram(fill = "#95D840FF") +
  labs(
    title = "Distribution of Wind Speed Values",
    x = "Wind Speed (miles per hour)"
  ) +
  theme(text = element_text(size = font_size))

rise_hist <- ggplot(data = data, aes(x = time_sun_rise)) +
  geom_histogram(bins = 15, fill = "#FDE725FF") +
  labs(
    title = "Distributions of Sunrise Times",
    x = "Hawaii Local Time"
  ) +
  theme(text = element_text(size = font_size))

set_hist <- ggplot(data = data, aes(x = time_sun_set)) +
  geom_histogram(bins = 15, fill = "#FDE725FF") +
  labs(
    title = "Distributions of Sunset Times",
    x = "Hawaii Local Time"
  ) +
  theme(text = element_text(size = font_size))

plot_grid(radiation_hist, temperature_hist, pressure_hist, humidity_hist, direction_hist, speed_hist, rise_hist, set_hist, ncol = 2)
```
From these histograms we make the following observations:

* Many values of `radiation` are either 0 or close to 0. These were probably observations taken at night. Using the `time`, `time_sun_rise`, and `time_sun_set` variables, we can make a new feature identifying whether an observation took place during the day or at night so our models will be able to separate out these values better.
* The `temperature`, `pressure`, and wind `speed` variables all seem to be roughly normally distributed. It appears that `pressure` is slightly skewed to the left, while `temperature` and wind `speed` are skewed to the right. I have no idea why the atmospheric pressure would be left-skewed, but this might have something to do with the local geography. The right skews on the temperature and wind speed make sense. Hawaii doesn't get very cold, but occasional heat waves could cause high temperatures. Wind speed can't go below 0, but it could be arbitrarily high (e.g. during a storm). 
* The `humidity` is heavily skewed towards 100%. That makes sense for Hawaii.
* The `wind-direction-degrees` histogram is difficult to interpret. The wind's direction is a continuous measure, but it doesn't make sense to record it as an angle clockwise from North. A NNE wind and a NNW wind come from practically the same direction, yet their degree measurements are very far apart: something like 30 degrees for NNE and 330 degrees for NNW. We will solve this problem by converting the wind direction to a factor variable with the levels North, East, South, and West, where each level covers 90 degrees of the circle, centered on the cardinal directions. There are other ways to solve this problem, but this solution is nice because it is simple to implement and easy to interpret.


### Data cleaning and feature engineering

#### Dealing with missing values

With this particular dataset, we got lucky that there were absolutely no missing values. Often, the data will have some `NA`s that we will have to deal with. Here's how we could deal with that in general:

* Look for `NA`s that mean something other than a lack of data, and replace them appropriately. Sometimes it might make sense to replace an `NA` in a numerical column with `0`, or to replace an `NA` in a character column with `"None"`.
* Impute missing numerical values using `method = "medianImpute"` in the `preProcess` function from the `caret` package. This will replace all `NA`s with the median of the existing values in their column.
* Impute missing categorical values by replacing `NA`s with the mode of the non-`NA` values in their column. Unfortunately, `preProcess` doesn't have such a method out of the box, so we would have to find another package that does this or write a function to do it ourselves.
* Alternatively, we could impute missing numerical and categorical variables using `method = "knnImpute"` in `preProcess`. This will replace the `NA`s with k-nearest neighbors predictions based on the rows that do not have `NA`s.

```{r, impute-missing, eval = FALSE}
# we were lucky to actually have no missing data
# but if we did, here's how we could medianImpute
data_clean <- data %>% preProcess(method = "medianImpute") %>%
  predict(newdata = data)
```

#### Feature engineering

In the previous section, we observed that we might benefit from creating a couple new variables. Let's do that now.

* `is_daytime` will be `0` if an observation occurred after sunrise and before sunset (the daytime) and `1` otherwise (the night).
* `wind_direction_factor` will be `"north"` if the wind angle was within 45 degrees of due North, `"east"` if within 45 degrees of due East, etc.

```{r, feature-engineering}
# create is_daytime boolean variable
data <- data %>% mutate(
  is_daytime = as.numeric((time > time_sun_rise) & (time < time_sun_set))
)

# create wind direction factor variable
data <- data %>% mutate(
  wind_direction_factor = case_when(
    wind_direction_degrees <= 45 ~ "north",
    wind_direction_degrees <= 135 ~ "east",
    wind_direction_degrees <= 225 ~ "south",
    wind_direction_degrees <= 315 ~ "west",
    wind_direction_degrees > 315 ~ "north"
  )
)
```

### A few more visualizations

Let's make a few more visuals based on our above observations, using our new features.

#### Radiation during the day and night

What is the daytime distribution of radiation? What is the nighttime distribution? Were we right in guessing that most of the near-0 values were observations occurring at night?

```{r, visualize-radiation}
# distribution of radiation values during the daytime
rad_day_hist <- ggplot(data = data %>% filter(is_daytime == 1), aes(x = radiation)) +
  geom_histogram(fill = "#F68F46FF") +
  labs(
    title = "Daytime Radiation",
    x = "Radiation (watts per square meter)"
  ) 

# distribution of radiation values during the night
rad_night_hist <- ggplot(data = data %>% filter(is_daytime == 0), aes(x = radiation)) +
  geom_histogram(fill = "#403891FF") +
  labs(
    title = "Nighttime Radiation",
    x = "Radiation (watts per square meter)"
  )

# distribution of daytime and nighttime radiation values on same plot
rad_group_hist <- ggplot(data = data, aes(x = radiation, group = factor(is_daytime))) +
  geom_histogram(
    position = "identity", 
    aes(fill = factor(is_daytime), color = factor(is_daytime)), 
    alpha = 0.6) +
  labs(
    title = "Radiation Grouped by Day/Night",
    x = "Radiation (watts per square meter)",
    fill = "",
    color = ""
  ) +
  scale_fill_viridis(discrete = TRUE, labels = c("Night", "Day"), option = "magma", begin = 0.2, end = 0.8) +
  scale_color_viridis(discrete = TRUE, labels = c("Night", "Day"), option = "magma", begin = 0.2, end = 0.8)

# distribution of radiation, colored by day/night
rad_fill_hist <- ggplot(data = data, aes(x = radiation)) +
  geom_histogram(aes(fill = factor(is_daytime))) +
  labs(
    title = "Radiation Filled by Day/Night",
    x = "Radiation (watts per square meter)",
    fill = ""
  ) +
  scale_fill_viridis(discrete = TRUE, labels = c("Night", "Day"), option = "magma", begin = 0.2, end = 0.8)

# show the plots
plot_grid(rad_day_hist, rad_night_hist, ncol = 2)
plot_grid(rad_group_hist, rad_fill_hist, nrow = 2)
```
Observations:

* Our assumption that most of the near-0 values for radiation occurred at night appears to be correct.
* There were still many near-0 values from the daytime. These values were likely observed at dawn or dusk when the daylight is not very bright, or on especially cloudy or stormy days when much of the sunlight gets blocked.
* There were some low non-zero values from the night. These values were likely observed at dawn or dusk when the night was just ending or starting, so there was still sunlight coming over the horizon. Or maybe if the moon is very bright, it will cause a small amount of observable solar radiation.

One more visual to track these observations:

```{r, visualize-radiation-2}
# scatterplot of radiation vs local time
ggplot(data = data, aes(x = time, y = radiation)) +
  geom_point(aes(color = factor(is_daytime)), alpha = 0.2) +
  labs(
    title = "Solar Radiation vs. Time of Day",
    x = "Hawaii Local Time",
    y = "Radiation (watts per square meter)",
    color = ""
  ) +
  scale_color_viridis(discrete = TRUE, labels = c("Night", "Day"), option = "magma", begin = 0.2, end = 0.8)
```
Looks like everything checks out. Because of the sharp turns in the data, it might be useful to fit a piecewise model like a spline.

#### Wind from the North, East, South, and West

In how many observations did the wind come from the North, East, South, and West? How did the average speed for each cardinal direction compare? What does the histogram of wind speeds look like when colored by direction?

```{r, visualize-wind}
# bar chart of number of observations of wind from each direction
wind_count_bar <- ggplot(data = data, aes(x = wind_direction_factor)) +
  geom_bar(stat = "count", aes(fill = wind_direction_factor)) +
  labs(
    title = "Count of Wind Oberservations for Each Cardinal Direction",
    x = "Wind Direction",
    y = "Number of Observations",
    fill = "Direction"
  ) +
  geom_text(aes(label = ..count..), stat = "count", vjust = -0.5, size = 3) +
  coord_cartesian(ylim = c(0, 16000)) +
  scale_fill_viridis(discrete = TRUE)

# bar chart of average speed of wind from each direction
wind_speed_bar <- ggplot(data = data, aes(x = wind_direction_factor, y = speed)) +
  geom_bar(stat = "summary", fun.y = "mean", aes(fill = wind_direction_factor)) +
  labs(
    title = "Average Wind Speed for Each Cardinal Direction",
    x = "Wind Direction",
    y = "Mean Wind Speed (mph)",
    fill = "Direction"
  ) +
  geom_text(aes(label = round(..y.., 4)), stat = "summary", fun.y = "mean", vjust = -0.5, size = 3) +
  coord_cartesian(ylim = c(0, 8)) +
  scale_fill_viridis(discrete = TRUE) 

# distribution of speed for eastern winds
speed_east_hist <- ggplot(
    data = data %>% filter(wind_direction_factor == "east"),
    aes(x = speed)) +
  geom_histogram(fill = "#440154FF", binwidth = 1.35) +
  labs(
    title = "Eastern Winds",
    x = "Wind Speed (miles per hour)"
  )

# distribution of speed for northern winds
speed_north_hist <- ggplot(
    data = data %>% filter(wind_direction_factor == "north"),
    aes(x = speed)) +
  geom_histogram(fill = "#33638DFF", binwidth = 1.35) +
  labs(
    title = "Northern Winds",
    x = "Wind Speed (mph)"
  )

# distribution of speed for southern winds
speed_south_hist <- ggplot(
    data = data %>% filter(wind_direction_factor == "south"),
    aes(x = speed)) +
  geom_histogram(fill = "#3CBB75FF", binwidth = 1.35) +
  labs(
    title = "Southern Winds",
    x = "Wind Speed (mph)"
  )

# distribution of speed for western winds
speed_west_hist <- ggplot(
    data = data %>% filter(wind_direction_factor == "west"),
    aes(x = speed)) +
  geom_histogram(fill = "#FDE725FF", binwidth = 1.35) +
  labs(
    title = "Western Winds",
    x = "Wind Speed (mph)"
  )

# distribution of eastern, northern, southern, and western wind speed on same plot
speed_group_hist <- ggplot(data = data, aes(x = speed, group = wind_direction_factor)) +
  geom_histogram(
    position = "identity", 
    aes(fill = wind_direction_factor, color = wind_direction_factor), 
    alpha = 0.4) +
  labs(
    title = "Wind Speed Grouped by Direction",
    x = "Wind Speed (mph)", 
    fill = "Direction",
    color = "Direction"
  ) +
  scale_fill_viridis(discrete = TRUE) +
  scale_color_viridis(discrete = TRUE)

# distribution of wind speed, colored by direction
speed_fill_hist <- ggplot(data = data, aes(x = speed)) +
  geom_histogram(aes(fill = wind_direction_factor)) +
  labs(
    title = "Wind Speed Filled by Direction",
    x = "Wind Speed (mph)",
    fill = "Direction"
  ) +
  scale_fill_viridis(discrete = TRUE)

# show the plots
plot_grid(wind_count_bar, wind_speed_bar, nrow = 2)
plot_grid(speed_east_hist, speed_north_hist, speed_south_hist, speed_west_hist, ncol = 2)
plot_grid(speed_group_hist, speed_fill_hist, nrow = 2)
```

Observations:

* The most common wind direction was South, and the least common was West.
* Western winds had the highest average speed, just slightly higher than the average speed for Southern winds. We don't know if any of the differences in means are statistically significant.
* The speeds for each direction appear to be approximately normally distributed. This means we could figure out whether the differences between the mean speeds are statistically significant by doing a two-sample T test on each pair of directions (but multiple tests increase the probability of a false positive). Alternatively, we could do an ANOVA test on all four directions (but if there is a significant difference between a pair of means, we won't know which pair).

#### Atmospheric pressure, temperature, and humidity

Is there a relationship between atmospheric pressure and temperature? What about between pressure and humidity?

```{r, visualize-pressure}
# scatterplot and simple regression of pressure vs temperature, colored by humidity
press_vs_temp <- ggplot(data = data, aes(x = temperature, y = pressure)) +
  geom_point(aes(color = humidity), alpha = 0.2) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Atmospheric Pressure vs. Temperature",
    x = "Temperature (Fahrenheit)",
    y = "Pressure (inHg)"
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "°")) +
  scale_color_viridis(option = "plasma", begin = 0.5, end = 1)

# scatterplot and simple regression of pressure vs humidity, colored by temperature
press_vs_hum <- ggplot(data = data, aes(x = humidity, y = pressure)) +
  geom_point(aes(color = temperature), alpha = 0.2) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Atmospheric Pressure vs. Humidity",
    x = "Humidity",
    y = "Pressure (inHg)"
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_color_viridis(option = "plasma")

# tiling of pressure on temperature-humidity grid
press_tile <- ggplot(
    data = data %>% group_by(temperature, humidity) %>% summarize(mean_pressure = mean(pressure)),
    aes(x = temperature, y = humidity)
  ) +
  geom_tile(aes(fill = mean_pressure)) +
  labs(
    title = "Pressure Tiling on Temperature and Humidity",
    x = "Temperature (Fahrenheit)",
    y = "Humidity",
    fill = "Mean Pressure"
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "°")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_viridis()

# show the plots
plot_grid(press_vs_temp, press_vs_hum, nrow = 2)
press_tile
```
Observations:

* There is a slight positive correlation between pressure and temperature. Even though the measurements are not from a closed system, this seems reasonable because the ideal gas law says $PV = nRT$, where $P$ is pressure and $T$ is temperature. As gas particles gain heat energy, their temperature increases and so does their speed. This makes the particles collide with things more frequently and faster, which we observe as increased pressure.
* There is a slight negative correlation between pressure and humidity. This makes sense because water vapor is less dense than dry air. When the amount of moisture in the air increases, the humidity increases and the density and pressure decrease.
* From the pressure tiling, it appears that in general lower pressures are associated with lower temperatures and higher humidities, while higher pressures are associated with higher temperatures and lower humidities. This tracks with our previous two observations.
* Obviously, we should **not** interpret the effects of temperature and humidity on pressure as causal. Weather systems are very complicated, so there are certainly variables not observed in our data that affect pressure, temperature, and humidity. Since we have not controlled for all confounders (and are not able to with this data), we cannot assume conditional independence. Thus, we cannot interpret the correlation as causal.

## Prediction models

We will try to predict the level of solar radiation from the other variables using three machine learning methods:

* K-Nearest Neighbors
* Elasticnet
* Random Forest.

Then we will evaluate the three models using the test RMSE (root mean squared error) to see which model did the best.

### Prepare data for model fitting

Before we can fit these models, we need to prepare the data.

* We will add an `id` column to number the rows, which will make them easy to keep track of.
* We already commented that `wind_direction_degrees` is a nonsensical/arbitrary way to measure wind direction, and similarly, `unix_time` is a nonsensical/arbitrary way to measure time. We won't use these variables in our models, so we can go ahead and drop them.
* R will interpret the data in `date` as categorical by default (because it is given as a character type), so we need to force it to numeric by reformatting the string and then calling the `lubridate` function `mdy()`. We also need to convert all dates and times to doubles so they can be standardized in `preProcess`.
* A random forest can use the data as-is, but KNN and elasticnet require the numerical predictors to be standardized (centered and scaled) and the categorical predictors to be converted to dummies. We will create a standardized version of the data for these models, and keep the original version for the random forest.
* We will hold out 20% of the data to evaluate model performance on unseen data at the very end.

```{r, adjust-columns}
# create an id column
data <- data %>% mutate(id = row_number()) %>% relocate(id)

# drop unix_time and wind_direction_degrees
data <- data %>% select(-unix_time, -wind_direction_degrees)

# reformat the date
data <- data %>% mutate(date = str_replace(date, " 12:00:00 AM", "")) %>% 
  mutate(date = mdy(date))

# convert all dates and times to doubles
data <- data %>% mutate(
  date = as.numeric(date),
  time = as.numeric(time),
  time_sun_rise = as.numeric(time_sun_rise),
  time_sun_set = as.numeric(time_sun_set)
)
```

```{r, standardize-data}
# create dummy variables
data_dummy <- dummyVars(
    radiation ~ .,
    data = data %>% select(-id),
    fullRank = TRUE
  ) %>% 
  predict(newdata = data) %>% 
  as.data.frame()

# standardize the non-boolean variables
# also don't standardize the id and outcome
data_std <- preProcess(
    x = data %>% select(-is_daytime, -id, -radiation),
    method = c("center", "scale")
  ) %>% 
  predict(newdata = data_dummy) %>% 
  as.data.frame()

# add back in the id and radiation columns
# which we took out in the processing of dummying and standardizing
data_std %<>% mutate(
    id = data$id,
    radiation = data$radiation
  ) %>%
  relocate(id)
```

```{r, split-data}
# set seed
set.seed(46848)

# randomly select 20% of the data to be used for testing
test_rows <- sample(
    x = data$id,
    size = nrow(data) * 0.2,
    replace = FALSE
  ) %>% as.numeric()

# split regular data
train <- data %>% filter(!id %in% test_rows)
test <- data %>% filter(id %in% test_rows) %>% select(-radiation)

# split standardized data
train_std <- data_std %>% filter(!id %in% test_rows)
test_std <- data_std %>% filter(id %in% test_rows) %>% select(-radiation)
```

### K-Nearest Neighbors

Explanation of KNN.

#### KNN Model 1: time, is_daytime, temperature

```{r, knn-model-1}
# set new seed for this code chunk
set.seed(98329)

# KNN model with only time, is_daytime, and temperature
knn_small <- train(
  # the model: predict radiation based on time, is_daytime and temperature
  radiation ~ time + is_daytime + temperature,
  data = train_std %>% select(-id),
  method = "knn",
  # tune parameters using 5-fold cross-validation
  trControl = trainControl(method = "cv", number = 5),
  # tuning parameter: the number of nearest neighbors, k
  tuneGrid = expand.grid(k = seq(1, 50, by = 1))
)
```

```{r, knn-model-1-viz}
knn_small_k <- ggplot(
    knn_small$results,
    aes(x = k, y = RMSE)
  ) +
  geom_line() +
  geom_point(aes(color = (k == knn_small[["bestTune"]][["k"]])), size = 2) +
  labs(
    title = "KNN Small: Tuning k (Number of Neighbors)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "k (Number of Neighbors)"
  ) +
  coord_cartesian(ylim = c(80, 180)) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")

knn_small_k
```

```{r, knn-model-1-predict}
# make vector of predictions
knn_small_pred <- knn_small %>% predict(newdata = test_std)

# make dataframe to compare with actual values
knn_small_df <- data.frame(
  id = test_std %>% select(id),
  radiation = data_std %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = knn_small_pred
)

head(knn_small_df)
```

#### KNN Model 2: all predictors

```{r, knn-model-2}
# set new seed for this code chunk
set.seed(71931)

# KNN model with only time, is_daytime, and temperature
knn_large <- train(
  # the model: predict radiation based on all other variables
  radiation ~ .,
  data = train_std %>% select(-id),
  method = "knn",
  # tune parameters using 5-fold cross-validation
  trControl = trainControl(method = "cv", number = 5),
  # tuning parameter: the number of nearest neighbors, k
  tuneGrid = expand.grid(k = seq(1, 50, by = 1))
)
```

```{r, knn-model-2-viz}
knn_large_k <- ggplot(
    knn_large$results,
    aes(x = k, y = RMSE)
  ) +
  geom_line() +
  geom_point(aes(color = (k == knn_large[["bestTune"]][["k"]])), size = 2) +
  labs(
    title = "KNN Large: Tuning k (Number of Neighbors)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "k (Number of Neighbors)"
  ) +
  coord_cartesian(ylim = c(80, 180)) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")

knn_large_k
```

```{r, knn-model-2-predict}
# make vector of predictions
knn_large_pred <- knn_large %>% predict(newdata = test_std)

# make dataframe to compare with actual values
knn_large_df <- data.frame(
  id = test_std %>% select(id),
  radiation = data_std %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = knn_large_pred
)

head(knn_large_df)
```

#### KNN Model 3: several variables (but not all)

```{r, knn-model-3}
# set new seed for this code chunk
set.seed(86129)

# KNN model with only time, is_daytime, and temperature
knn_med <- train(
  # the model: predict radiation based several variables
  radiation ~ time + is_daytime + temperature + date + pressure + humidity,
  data = train_std %>% select(-id),
  method = "knn",
  # tune parameters using 5-fold cross-validation
  trControl = trainControl(method = "cv", number = 5),
  # tuning parameter: the number of nearest neighbors, k
  tuneGrid = expand.grid(k = seq(1, 50, by = 1))
)
```

```{r, knn-model-3-viz}
knn_med_k <- ggplot(
    knn_med$results,
    aes(x = k, y = RMSE)
  ) +
  geom_line() +
  geom_point(aes(color = (k == knn_med[["bestTune"]][["k"]])), size = 2) +
  labs(
    title = "KNN Medium: Tuning k (Number of Neighbors)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "k (Number of Neighbors)"
  ) +
  coord_cartesian(ylim = c(80, 180)) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")

knn_med_k
```

```{r, knn-model-3-predict}
# make vector of predictions
knn_med_pred <- knn_med %>% predict(newdata = test_std)

# make dataframe to compare with actual values
knn_med_df <- data.frame(
  id = test_std %>% select(id),
  radiation = data_std %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = knn_med_pred
)

head(knn_med_df)
```
#### KNN Model 4: same as Model 3 but without is_daytime

```{r, knn-model-4}
# set new seed for this code chunk
set.seed(38473)

# KNN model with only time, is_daytime, and temperature
knn_med2 <- train(
  # the model: predict radiation based on is_daytime and temperature
  radiation ~ time + temperature + date + pressure + humidity,
  data = train_std %>% select(-id),
  method = "knn",
  # tune parameters using 5-fold cross-validation
  trControl = trainControl(method = "cv", number = 5),
  # tuning parameter: the number of nearest neighbors, k
  tuneGrid = expand.grid(k = seq(1, 50, by = 1))
)
```

```{r, knn-model-4-viz}
knn_med2_k <- ggplot(
    knn_med2$results,
    aes(x = k, y = RMSE)
  ) +
  geom_line() +
  geom_point(aes(color = (k == knn_med2[["bestTune"]][["k"]])), size = 2) +
  labs(
    title = "KNN Medium 2: Tuning k (Number of Neighbors)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "k (Number of Neighbors)"
  ) +
  coord_cartesian(ylim = c(80, 180)) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")

knn_med2_k
```

```{r, knn-model-4-predict}
# make vector of predictions
knn_med2_pred <- knn_med2 %>% predict(newdata = test_std)

# make dataframe to compare with actual values
knn_med2_df <- data.frame(
  id = test_std %>% select(id),
  radiation = data_std %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = knn_med2_pred
)

head(knn_med2_df)
```

```{r, knn-all-viz}
# see charts for all knn models together
plot_grid(knn_small_k, knn_large_k, knn_med_k, knn_med2_k, nrow = 2)
```

### Elasticnet

Explanation of Elasticnet = ridge + lasso.

```{r, elasticnet-model}
# set a new seed for this chunk
set.seed(83352)

# set range of lambdas, the size of the penalty
#lambdas = 10 ^ seq(from = 1, to = -3, length = 1e2)
lambdas = seq(from = 0, to = 4, by = 0.1)

# set range of alphas, the relative weight on ridge (0) and lasso (1) penalties
alphas = seq(from = 0, to = 1, by = 0.05)

# elasticnet model
elasticnet <- train(
  # the model: regress radiation on all predictors
  radiation ~ .,
  data = train_std %>% select(-id),
  method = "glmnet",
  # evaluate performance with 5-fold cross validation
  trControl = trainControl(method = "cv", number = 5),
  # the tuning parameters: the alphas and lambdas defined above
  tuneGrid = expand.grid(alpha = alphas, lambda = lambdas)
)

```

```{r, elasticnet-model-predict}
# make vector of predictions
elasticnet_pred <- elasticnet %>% predict(newdata = test_std)

# make dataframe to compare with actual values
elasticnet_df <- data.frame(
  id = test_std %>% select(id),
  radiation = data_std %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = elasticnet_pred
)

head(elasticnet_df)
```

```{r, elasticnet-model-viz}
elasticnet_tile <- ggplot(
    data = elasticnet$results, #%>% filter(alpha > 0),
    aes(x = alpha, y = lambda)
  ) +
  geom_tile(aes(fill = RMSE)) +
  labs(
    title = "Elasticnet: Tuning α and λ",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "α (0 = ridge, 1 = lasso)",
    y = "λ (shrinkage penalty)"
  ) +
  scale_fill_viridis(option = "plasma", begin = 0, end = 1)

elasticnet_tile2 <- ggplot(
    data = elasticnet$results %>% filter(alpha > 0),
    aes(x = alpha, y = lambda)
  ) +
  geom_tile(aes(fill = RMSE)) +
  labs(
    title = "Elasticnet: Tuning α and λ",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "α (0 = ridge, 1 = lasso)",
    y = "λ (shrinkage penalty)"
  ) +
  scale_fill_viridis(option = "plasma", begin = 0, end = 1)

plot_grid(elasticnet_tile, elasticnet_tile2)
```

### Tree-based methods

#### Decision tree

```{r, tree-model}
set.seed(64395)

tree <- train(
  radiation ~ .,
  data = train %>% select(-id),
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = data.frame(cp = seq(0, 0.2, by = 0.01))
)
```

```{r, tree-model-predict}
# Make predictions on test set
tree_pred = tree %>% predict(newdata = test)

# The results
tree_df <- data.frame(
  id = test %>% select(id),
  radiation = data %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = tree_pred
)

head(tree_df)
```

```{r, tree-model-viz}
ggplot(
  data = tree$results,
  aes(x = cp, y = RMSE)
) +
  geom_line() +
  geom_point(aes(color = (cp == tree[["bestTune"]][["cp"]]))) +
  labs(
    title = "Decision Tree: Tuning cp (Complexity Parameter)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "cp (Complexity Parameter)"
  ) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")
```
```{r, tree-model-2}
set.seed(64395)

tree_small <- train(
  radiation ~ .,
  data = train %>% select(is_daytime, temperature, radiation),
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = data.frame(cp = seq(0.01, 0.2, by = 0.01))
)
```

```{r, tree-model-2-predict}
# Make predictions on test set
tree_small_pred = tree_small %>% predict(newdata = test)

# The results
tree_small_df <- data.frame(
  id = test %>% select(id),
  radiation = data %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = tree_small_pred
)

head(tree_small_df)
```

```{r, tree-model-2-viz}
ggplot(
  data = tree_small$results,
  aes(x = cp, y = RMSE)
) +
  geom_line() +
  geom_point(aes(color = (cp == tree_small[["bestTune"]][["cp"]]))) +
  labs(
    title = "Decision Tree: Tuning cp (Complexity Parameter)",
    subtitle = "Using 5-fold CV to minimize RMSE",
    x = "cp (Complexity Parameter)"
  ) +
  scale_color_manual(values = c("#314f4f", "#e64173")) +
  theme(legend.position = "none")
```

```{r, tree-model-viz-2}
rpart.plot(
  x = tree_small$finalModel,
  type = 4,
  extra = "auto",
  box.palette = "Oranges",
  branch.lty = 3,
  shadow.col = "gray",
  nn = TRUE,
  digits = 3
)
```

#### Random forest

Explanation of Random Forests (how you get them from Random Trees)

```{r, forest-model}
# set a new seed for this chunk
set.seed(42712)

# random forest model
forest = train(
  # The model: predict radiation based on everything else
  radiation ~ .,
  # The data: non-pre-processed train dataframe, without id
  data = train %>% select(-id),
  # Use ranger to implement random forest with 100 trees
  method = "ranger",
  num.trees = 100,
  # Evaluate performance with out-of-bag error estimation
  trControl = trainControl(method = "oob"),
  # Tuning parameters: 
    # mtry - number of random predictors at each split,
    # splitrule - the rule for splitting,
    # and min.node.size - minimum number of observations per leaf
  tuneGrid = expand.grid(
    "mtry" = c(1, 2, 3, 4, 5, 6, 7, 8),
    "splitrule" = "variance",
    "min.node.size" = 1:10
  )
)
```


```{r, forest-model-predict}
# Make predictions on test set
forest_pred = forest %>% predict(newdata = test)

# The results
forest_df <- data.frame(
  id = test %>% select(id),
  radiation = data %>% 
    filter(id %in% test_rows) %>% 
    select(radiation),
  pred = forest_pred
)

head(forest_df)
```

### Evaluation of models

How well did our models work?

```{r, evaluation}
# Define a function that takes in our results dataframes and returns test RMSE
rmse <- function(df){
    (df$radiation - df$pred)^2 %>% mean() %>% sqrt()
}

# Find the RMSE for each model
all_data <- data.frame(
  model_name = c("knn_small", "knn_large", "knn_med", "knn_med2", "elasticnet", "tree_small", "tree", "forest"),
  rmse_vals = c(rmse(knn_small_df), rmse(knn_large_df), rmse(knn_med_df), rmse(knn_med2_df), rmse(elasticnet_df), rmse(tree_small_df), rmse(tree_df), rmse(forest_df))
)
```

```{r, evaluation-viz}
ggplot(all_data, aes(x = model_name, y = rmse_vals)) +
  geom_col(aes(fill = rmse_vals)) +
  geom_text(aes(label = round(rmse_vals, 4)), vjust = -0.5) +
  labs(
    title = "Test RMSE Values for Our Models",
    x = "Model Name",
    y = "Test RMSE",
    fill = "Test RMSE"
  ) +
  scale_fill_viridis(
    discrete = FALSE,
  ) +
  coord_cartesian(ylim = c(0, 200))

```

## Conclusion

The standard deviation of radiation values in the data is `r sd(data$radiation)`. Our best model was the random forest with a test RMSE of `r rmse(forest_df)`, much lower than the standard deviation. The mid-sized KNN model also did well, with a test RMSE of `r rmse(knn_med_df)`. This suggests that we can predict solar radiation pretty well using other meteorological variables.

